Calendar-based graphics for visualising people's daily schedules

### 2017-07-15 main.Rmd
Added one more paragraph to intro and referenced to geofacet

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time}, which is complementary to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

1. Variations primarily result from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurred events).
2. Since the data are often collected at daily frequency or a time scale more frequent than daily, they typically involve a large number of observations spanning over a long time period.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need for comparing and contrasting between locations.

```{r ped-map, fig.width = 4, out.width = "0.55\\linewidth", fig.cap = "Map of the Melbourne city along with purple dots indicating sensor locations."}
```

@@

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using coloured blocks rather than curves, it may become perceptually difficult to accurately judge the shape positions and changes, although the glyphs come with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout, (3) to enable an overlaying plot consisting of multiple time series. The proposed algorithm has been implemented as the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

@@

Section \ref{sec:algorithm} discusses the implementation applied to the fixed scale which is using the whole range of all the data values. The \code{scale} argument controls the scaling of the display. In addition to the fixed scale (\code{fixed}), the remaining options include free scale on all the days (\code{free}), day of the week only (\code{free_wday}), and day of the month only (\code{free_mday}) for different comparisons.

@@

```{r write-bib}
write_bib(c(.packages()), 'packages.bib')
```

>>>>

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time} provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

1. Variations primarily result from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurred events).
<!-- 2. Temporal data plots should be arranged in a way that serves a familiar format to viewers. -->
2. Since the data are often collected at daily frequency or a time scale more frequent than daily, they typically involve a large number of observations spanning over a long time period.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need in comparing and contrasting between locations.

```{r ped-map, fig.width = 4, out.width = "0.55\\linewidth", fig.cap = "Map of the Melbourne city with purple dots indicating sensor locations."}
```

@@

A collection of data plots, organised in a familiar format, offer a useful and intuitive way to tell a richer story and reveal more complex relationships. @Wickham2012glyph embedded a sequence of daily temperature line charts into a glyph map according to the spatial locations; and @R-geofacet provided methods to arrange the data plots of many types into a preserved geographical grid. They both attempt to have the self-contained data plots in spatial context.

Alternatively, calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using coloured blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using the curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout, (3) to enable an overlaying plot consisting of multiple time series. The proposed algorithm has been implemented to the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

@@

Section \ref{sec:algorithm} discusses the implementation applied to the fixed scale which is using the whole range of all the data values. The \code{scale} argument controls the scaling of the display. In addition to the fixed scale (\code{fixed}), the remaining options include free scale on all the days (\code{free}), day of the week only (\code{free_wday}), and day of the month only (\code{free_mday}) for different comparisons and emphases.

@@

```{r write-bib}
write_bib(c(.packages(), "geofacet"), 'packages.bib')
```

### 2017-07-15 main.tex
Added one more paragraph to intro and referenced to geofacet

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 3 sensors. Figure \ref{fig:facet-time}, which is
complementary to the Figure \ref{fig:time-series-plot}, provides another
aspect of temporal patterns in the more detailed level by faceting day
of the week. It lends itself to a number of issues that make exploratory
data visualisation challenging in many temporal-context applications
involving human behaviours:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily result from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurred events).
\item
  Since the data are often collected at daily frequency or a time scale
  more frequent than daily, they typically involve a large number of
  observations spanning over a long time period.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need for comparing and contrasting
  between locations.
\end{enumerate}

@@

\caption[Map of the Melbourne city along with purple dots indicating sensor locations]{Map of the Melbourne city along with purple dots indicating sensor locations.}\label{fig:ped-map}
\end{figure}
\end{CodeChunk}

@@

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week as the smallest time scale. Time of day, which serves as one of the
most important aspect in explaining variations arisen from pedestrian
sensor data, will be neglected through daily aggregation. Additionally
if simply using coloured blocks rather than curves, it may become
perceptually difficult to accurately judge the shape positions and
changes, although the glyphs come with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to
incorporate line graphs and other types of glyphs into the graphical
toolkit for the calendar layout, (3) to enable an overlaying plot
consisting of multiple time series. The proposed algorithm has been
implemented as the \code{frame_calendar} function in the \pkg{sugrrants}
package \citep{R-sugrrants} using \proglang{R} \citep{R-base}.

>>>>

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. A small multiples, shown in Figure
\ref{fig:time-series-plot}, gives an overall picture of foot traffic at
a selection of 3 sensors. Figure \ref{fig:facet-time} provides another
aspect of temporal patterns in the more detailed level by faceting day
of the week. It lends itself to a number of issues that make exploratory
data visualisation challenging in many temporal-context applications
involving human behaviours:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily result from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurred events). 
\item
  Since the data are often collected at daily frequency or a time scale
  more frequent than daily, they typically involve a large number of
  observations spanning over a long time period.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need in comparing and contrasting
  between locations.
\end{enumerate}

@@

\caption[Map of the Melbourne city with purple dots indicating sensor locations]{Map of the Melbourne city with purple dots indicating sensor locations.}\label{fig:ped-map}
\end{figure}
\end{CodeChunk}

@@

A collection of data plots, organised in a familiar format, offer a
useful and intuitive way to tell a richer story and reveal more complex
relationships. \citet{Wickham2012glyph} embedded a sequence of daily
temperature line charts into a glyph map according to the spatial
locations; and \citet{R-geofacet} provided methods to arrange the data
plots of many types into a preserved geographical grid. They both
attempt to have the self-contained data plots in spatial context.

Alternatively, calendar-based graphics turn out to be a useful tool in
unfolding human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
weekdays and weekends, highlights public holiday, and presents other
known seasons like school vacations, all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in a couple of R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are too
constrained to colour-encoding graphics and day of week as the smallest
time scale. Time of day, which serves as one of the most important
aspect in explaining variations arisen from pedestrian sensor data, will
be neglected through daily aggregation. Additionally if simply using
coloured blocks rather than curves, it may become perceptually difficult
to estimate the shape positions and changes, although using the curves
comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to
incorporate line graphs and other types of glyphs into the graphical
toolkit for the calendar layout, (3) to enable an overlaying plot
consisting of multiple time series. The proposed algorithm has been
implemented to the \code{frame_calendar} function in the \pkg{sugrrants}
package \citep{R-sugrrants} using \proglang{R} \citep{R-base}.
